### 2013-02-13 doc/xps2013/proposal.tex
HW support for aggregation. added sections collaboration plan and data management

In this proposal we first describe our early
Grappa~\cite{grappa} prototype, a runtime system for irregular
applications. It provides several facilities: a global shared memory
abstraction, user-level threading support, and communication and
synchronization primitives. Applications targeted for Grappa are not expected
to have \emph{any} locality (although what locality that exists is favorably
exploited), but are expected to provide many thousands of concurrent threads.

@@

Large data that is expected to be shared and accessed with low locality is stored in Grappa's global memory. All global data must be accessed through calls into Grappa's API, shown in Figure~\ref{fig:accessing-memory}.

@@

% Should we discuss doing HW support for aggregation? Using NetFPGAs?

@@

\paragraph{Technology transfer:} Bill Dally during his 2010 Eckert-Mauchly award speech said (in effect), technology transfers happen when people transfer.  Having just come back from a three year stint founding a startup company (where we transferred the Deterministic Multiprocessing technology from UW), we've seen this first hand.  The students and faculty involved in the Grappa project will spin off a company with their work.  The exact time frame is not yet known, but everyone involved in the project is excited to see the technology underpin a startup company and make this technology available to the broader commercial world.

@@

%% TODO: need to expand below.

\noindent\textbf{Year 1:} The first year of this effort will focus on creating the benchmark suite and initial language and compiler support.  We aim to pull together 6-8 benchmarks, sample data sets, and scripts to measure performance results on the initial Grappa runtime, UPC~\cite{upc} implementations, and serial reference code.  For the compiler we will have the front end done enabling easy addition of keywords to annote data-types and structural elements (parallel for, etc).  The compiler will be generating in-efficient Grappa code by the end of the first year.

@@

>>>>

In this proposal we first describe our early
Grappa prototype, a runtime system for irregular
applications. It provides several facilities: a global shared memory
abstraction, user-level threading support, and communication and
synchronization primitives. Applications targeted for Grappa are not expected
to have \emph{any} locality (although what locality that exists is favorably
exploited), but are expected to provide many thousands of concurrent threads.

@@

Large data that is expected to be shared and accessed with low locality is stored in Grappa's global memory. All global data must be accessed through calls into Grappa's API. %, shown in Figure~\ref{fig:accessing-memory}.

@@

\paragraph{Hardware support for aggregation:} If the increased latency caused
by extra hops in hierarchical aggregation turns out to be a big factor, we
will explore reducing by move some of the aggregation logic to the network
card, either by exploiting off-the-shelf programmable network cards or
experimental systems such ask NetFPGA~\cite{netfpga} (while we are planning to
focus on Infiniband-based networking, if we explore the NetFPGA route we will
be using Ethernet).

@@

\paragraph{Technology transfer:} Bill Dally during his 2010 Eckert-Mauchly award speech said (in effect), technology transfers happen when people transfer.  Having just come back from a three year stint founding a startup company (where we transferred the Deterministic Multiprocessing technology from UW), we've seen this first hand.  The students and faculty involved in the Grappa project plan to spin off a company with their work.  The exact time frame is not yet known, but everyone involved in the project is excited to see the technology underpin a startup company and make this technology available to the broader commercial world.

@@

\noindent\textbf{Year 1:} The first year of this effort will focus on creating the benchmark suite and initial language and compiler support.  We aim to pull together 6-8 benchmarks, sample data sets, and scripts to measure performance results on the initial Grappa runtime, UPC~\cite{upc} implementations, and serial reference code.  For the compiler we will have the front end done enabling easy addition of keywords to annotate data-types and structural elements (parallel for, etc).  The compiler will be generating in-efficient Grappa code by the end of the first year.

@@

\newpage

\section*{Data Management Plan}

The proposed research involves developing experimental systems and
assessing how well they work. The kinds of data we will produce
include: characterization data of program behavior; data on system
performance and reliability; data on how programmers use new language
constructs and tools. As is our practice, we will regularly submit
papers for publication detailing our findings and design of our
systems.

As we have already been doing with all our research, for every
published paper, we make all the raw data available online. This
allows other researchers to look at the data in different ways, and it
enables them to reproduce results. We also often publish
the experimental software developed, following the rules outlined by
the University intellectual property office --- which provides ample
rights for other researchers to build upon our experimental
systems. We do not anticipate using privileged or confidential
information.

\newpage
\section*{Collaboration Plan}
%\setcounter{page}{1}

Because the PIs already successfully collaborate and co-advise an
active cadre of spectacular students, the necessary tools and
coordination mechanisms are already in place for this project to
proceed successfully.  We first describe the PIs' complementary
backgrounds, then describe the current organization of our (amorphous)
research group, then describe the specific technology and facilities
that enable seamless coordination, and finally describe specific PI roles
for the research in this proposal.

\subsection*{Complementary PI Experience}

PI Oskin has traditionally participated in the computer-architecture research
community, co-PI Ceze, in addition to the architecture community also has a
presence in the systems and programming languages community, and co-PI Kahan
has participated mainly in the parallel and scientific computing community.
The move toward multicore architectures has re-invigorated collaboration
across these divides and recent joint work by Oskin, Ceze, and Kahan
embodies this important trend. There are currently five graduate students for
which at last two of the PIs serve as advisors. In addition to the project
described in this proposal, they are also working together on techniques for
detecting concurrency errors with machine learning and language
specifications, using disciplined approximations to save energy in computing,
and developing new operating-system abstractions for emerging non-volatile
memory devices. What has made these collaborations so fruitful in such a short
period of time is the PIs' complementary backgrounds. We now briefly describe
these backgrounds.

Oskin has... {\bf MARK PUT YOUR STUFF HERE.}

Ceze has been conducting research on architecture support, compiler
techniques, and programming models to improve the programmability of
multiprocessor systems for seven years and has co-authored over thirty
papers in these areas.  His research includes advances in
architecture~\cite{tlsooo, bulk, bulksc, swbulk, cyclops, dmp,
  delorean} and programming models~\cite{colorama, ipot, posh} to
simplify exploiting concurrency in multiprocessors. Some of his recent
results especially related to this proposal include a programming
model based on ordered transactions and data annotations~\cite{ipot};
improving reliability of multithreaded software with architecture
support for bug detection and avoidance~\cite{atomaid, aatoppicks,
  bugaboo, oshajava, cs-isca10, ce-isca10}; and deterministic
execution of multithreaded programs~\cite{dmp, dmptoppicks,
  asplos10coredet, dmpos}. PI Ceze recently co-organized an NSF-funded
workshop focused specifically on deterministic multiprocessing. The
workshop was very successful, bringing leaders in the multicore area
from industry and academia.

Kahan has tremendous experience in ... from Tera/Cray ... {\bf SIMON PUT YOUR STUFF HERE.}

\subsection*{The SAMPA Research Group: Addressing the Concurrency
  Challenge Across the Execution Stack}

The PIs organize their joint research activity under the umbrella of
the ``SAMPA group'' (SAfe MultiProcessing Architectures).  Some brief
statistics help summarize the group's size, strength, and focus:
\begin{itemize}
\item There are roughly ten graduate students, most actively
  coadvised.
\item In the last four years, the group has produced twelve papers
  published in ASPLOS, ISCA, PLDI, OOPSLA, MICRO, and OSDI as well as three
  papers in the USENIX Workshop on Hot Topics in Parallelism.
\end{itemize}
The SAMPA group is an ideal size for collaboration across the system
execution stack from computer architecture up through software
engineering.  Compared to large collaborations with several faculty,
there are no graduate students (or faculty) who specialize entirely in
particular subtopics.  The students gain expertise across
computer-science disciplines and use appropriate ideas from hardware,
software, theory, etc. to solve research problems.

\subsection*{Specific Technology and Facilities}

The research group that will perform the research has the facilities
needed for successful collaboration.  There is a shared lab facility
where weekly meetings are held.  Students in the group share computing
infrastructure, including dedicated machines and regular use of cloud
services such as Amazon's EC2 for scalability experiments.

The PIs regularly hold joint meetings with each student that they
co-advise.  They do not split advising duties so much as duplicate
them.

\subsection*{Specific Roles and Responsibilities}

It should be clear from the preceding discussions that the PIs are
entirely comfortable jointly managing the work in the proposal and the
students involved.  Nonetheless, several specific pieces of the work
more naturally align with the expertise and interests of one of the
PIs:

\begin{itemize}
\item PI Oskin will focus on communication issues. 
\item PI Ceze will focus on compiler and programming languages work.
\item PI Kahan will focus on organizing benchmarks and performance evaluation efforts.
\item All PIs expect to solicit guidance and collaboration from
  researchers such as John Feo (PNNL), Jim Larus (Microsoft Research), Hans
  Boehm (HP Research), and Preston Briggs,  with whom
  they have already collaborated on other projects.
\end{itemize}
